---
title: "Using Example FlowJo v11 Data"
author: "CyFj11 Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Example FlowJo v11 Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Using Example FlowJo v11 Data

This document demonstrates how to use the example FlowJo v11 data included with the package for testing and development purposes.

## Loading the Example Data

The package includes an example FlowJo v11 workspace file that can be used for testing:

```{r load-example}
library(CyFj11)

# Load the example workspace using the helper function
workspace <- load_example_workspace()

# Examine the structure
str(workspace, max.level = 2)
```

## Exploring the Workspace Contents

Let's look at what's in the workspace:

```{r explore-workspace}
# Basic information
cat("Workspace path:", workspace$path, "\n")
cat("Number of groups:", length(workspace$groups), "\n")
cat("Number of data sources:", length(workspace$dataSources), "\n")
cat("Number of population definitions:", length(workspace$populationDefinitions), "\n")
cat("Number of populations:", length(workspace$populations), "\n")
```

## Working with Gates

Let's examine some of the gates in the workspace:

```{r examine-gates}
# Look at the population definitions
pop_names <- sapply(workspace$populationDefinitions, function(pop) {
  pop$definition$name
})

# Show first 10 population names
head(pop_names, 10)

# Look at a specific population definition
first_pop_uuid <- names(workspace$populationDefinitions)[1]
first_pop <- workspace$populationDefinitions[[first_pop_uuid]]
first_pop$definition
```

## Using in Your Own Tests

You can use this example data in your own tests. Here's a template:

```{r test-template, eval=FALSE}
test_that("my function works with FlowJo data", {
  # Load the example workspace
  workspace <- load_example_workspace()
  
  # Test your function
  result <- my_analysis_function(workspace)
  
  # Make assertions
  expect_s3_class(result, "my_result_class")
  expect_gt(length(result$gates), 0)
})
```

## Direct File Access

You can also access the file directly if needed:

```{r direct-access, eval=FALSE}
# Get the path to the example file
test_file <- system.file("extdata", "test.data.flowjo", package = "CyFj11")

# Load it directly
workspace <- read_flowjo11_workspace(test_file)
```

This approach is useful when you need to test error conditions or specific file handling scenarios.

## Conclusion

The example FlowJo v11 data provides a real-world dataset for testing and development. It includes a complete workspace with groups, data sources, and population definitions that represent a typical cytometry analysis workflow.