---
title: "EXPERIMENT_3_COMPATIBILITY_PROTOCOL"
output:
  html_document: default
---
# Experiment 3: FlowJo v11→v10 Backward Compatibility Protocol

**Objective:** Provide quantitative evidence that FlowJo v11 can import v10 workspaces without information loss for features used in CyFj11 conversion.

=========

Version 11.1.0 (11.1.0.178992)
Engine Version: 42.18.10
Engine Build Time: Fri Oct 31 16:13:24 2025
Engine Commit Hash: 97e4b65b4

Transformations (FLv10) -> import to FL11 
linear -> linear
log -> biexp
biex -> biexp
arcSinh -> linear
hyperlog -> biexp
logicle -> linear
miltenyi - > biexp

Transformations (FLv10) -> import to CytoML/gatingset 
linear -> linear
log -> log
biex -> biexp
arcSinh -> fasinh
hyperlog -> ERORR
logicle -> biexp
miltenyi - > ERROR

Transformation (Fl11) -> CyFl11
linear -> linear
log -> log10
biex -> biex
arcTanh (not working) -> NA

Gates:
Fl11 -> CyFl11
Elipse
poligon
rectangle
range/histogram
quadrant
logical

Gates:
FL10 -> FL11
to test:
Elipse
poligon
rectangle
range/histogram
quadrant
logical



Gates:
FL10 -> CytoML/gatingset
to test:
Elipse 
poligon
rectangle
range/histogram
quadrant
logical


```{r setup}
library(CytoML)
library(flowWorkspace)
library(ggcyto)
library(tidyverse)    # includes ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats
library(janitor)
library(gridExtra)
library(knitr)
library(kableExtra)
library(GGally)
library(patchwork)
source("pretty_print_flowjo.R")

# Read the file - readr handles the complex headers better than base R
statsFj11 <- read_csv("data/Statistics.fj11.csv", 
               name_repair = "unique",  # Ensures unique column names
               na = c("", "NA", "−1.00"))  # Handle the unicode minus in missing values
statsFj11 = statsFj11[c(1,2),]

# Read the file - readr handles the complex headers better than base R
statsFj10 <- read_csv("data/statsFL10.csv", 
               name_repair = "unique",  # Ensures unique column names
               na = c("", "NA", "−1.00"))  # Handle the unicode minus in missing values
statsFj10 = statsFj10[c(1,2),]

colnames(statsFj10)
colnames(statsFj11)

```


```{r}

# Function to create standardized short names with case preserved
standardize_column_names <- function(colnames_vector) {
  
  standardized <- sapply(colnames_vector, function(name) {
    
    # Handle special cases first
    if (grepl("^File Name$|^\\.\\.\\.1$", name)) {
      return("sample_id")
    }
    
    if (grepl("^Ungated Count$|^Count$", name) && !grepl("/|\\|", name)) {
      return("ungated_cnt")
    }
    
    # Extract population path
    # Remove "Ungated / " prefix if present
    clean_name <- gsub("^Ungated / ", "", name)
    
    # Determine if it's frequency or count
    is_freq <- grepl("Freq\\. of Parent", name)
    is_count <- grepl("Count", name)
    
    # Extract just the population path (before " | " or " Freq" or " Count")
    pop_path <- clean_name %>%
      # Remove the metric part
      gsub(" \\| Freq\\. of Parent \\(%\\)", "", .) %>%
      gsub(" \\| Count", "", .) %>%
      gsub(" Freq\\. of Parent", "", .) %>%
      gsub(" Count", "", .) %>%
      # Clean up path separators
      gsub(" / ", "/", .) %>%
      # Remove common prefix that doesn't distinguish populations
      gsub("^cells/Singlets/Single_cells/", "", .) %>%
      gsub("^cells/Singlets/", "", .) %>%
      gsub("^cells/", "", .) %>%
      # Replace "/" with "_"
      gsub("/", "_", .) %>%
      # Handle special characters
      gsub("\\+\\-", "pm", .) %>%  # MHCII+- becomes MHCIIpm
      gsub("\\+", "pos", .) %>%     # CD14+ becomes CD14pos, MHCII+ becomes MHCIIpos
      gsub("\\-", "neg", .) %>%     # CD14- becomes CD14neg
      gsub("Q\\d+: ", "", .) %>%    # Remove "Q5: ", "Q6: ", etc.
      gsub(":", "", .) %>%           # Remove remaining colons
      gsub(",", "", .) %>%           # Remove commas
      gsub(" ", "_", .) %>%          # Spaces to underscores
      # Clean up multiple underscores
      gsub("_+", "_", .) %>%
      # Remove leading/trailing underscores
      gsub("^_|_$", "", .)
    
    # Add suffix (keep case as-is, don't convert to lowercase)
    if (is_freq) {
      return(paste0(pop_path, "_frq"))
    } else if (is_count) {
      return(paste0(pop_path, "_cnt"))
    } else {
      return(pop_path)
    }
    
  }, USE.NAMES = FALSE)
  
  return(standardized)
}

```


## Experimental Design

### Phase 1: Create Test Workspaces in FlowJo v11

**You will need:**
- FlowJo v11 (any recent version)
- Sample FCS files (we can provide these)

**Test Cases to Create:**

| Test Case | Description | Gates to Include |
|-----------|-------------|------------------|
| TC-1 | Simple rectangular | 2-3 rectangle gates on single sample |
| TC-2 | Polygon gates | 2-3 polygon gates with 4-8 vertices |
| TC-3 | Ellipsoid gates | 1-2 ellipsoid gates |
| TC-4 | Boolean combination | AND, OR, NOT operations (≥3 parent populations) |
| TC-5 | Nested hierarchy | 3-level gating tree (parent → child → grandchild) |
| TC-6 | Quadrant gates | Standard 4-quadrant split |
| TC-7 | Multiple samples | 5-10 samples with identical gating |
| TC-8 | With compensation | Spillover matrix applied |
| TC-9 | With transformations | Biexponential on both axes |
| TC-10 | Complex real-world | 10+ gates, mixed types, 3+ samples |

MHCII+ (7.86%)
├── Q5: CD14+ CD11c- (2.00%) → Classical monocytes
├── Q6: CD14+ CD11c+ (3.66%) → Intermediate monocytes  
├── Q7: CD14- CD11c+ (9.95%) → mDCs
└── Q8: CD14- CD11c- (84.4%) → Other (pDCs, lymphocytes)

To demonstrate the analytical pipeline for flow cytometry data processing, we applied a sequential gating strategy to identify major innate immune cell populations and their cytokine production profiles. Briefly, single cells were identified using standard FSC-A/SSC-A gating to exclude debris (68.6% of events), followed by doublet discrimination using FSC-A/FSC-H (99.8% singlets) and SSC-A/SSC-H gates (99.3% single cells). From single cells, we separated antigen-presenting cells (APCs) as MHCII+ (7.86%) and MHCII- populations (92.1%). Within the MHCII+ population, we applied a quadrant gate (CD14 vs CD11c) to identify classical monocytes (Q5: CD14+ CD11c-, 2.00%), intermediate monocytes (Q6: CD14+ CD11c+, 3.66%), myeloid dendritic cells (mDCs: CD14- CD11c+, 9.95%), and other APCs (Q8: CD14- CD11c-, 84.4%). Total monocytes (5.65%) were defined as the combined Q5 and Q6 populations using a logical OR gate. Within the MHCII- population, the same quadrant gating identified neutrophils (CD14dim CD11c+/-, 18.5%) and lymphocytes (Q8: CD14- CD11c-, 81.0%). Functional assessment included cytokine production gates for TNFα, IL-6, and IL-12 in monocytes, mDCs, and neutrophils. This gating strategy is presented for demonstration purposes to illustrate the computational workflow and does not represent the primary analysis of the manuscript.


dataDir <- system.file("extdata", package = "flowWorkspaceData")
dir(path=dataDir)
file.copy(file.path(dataDir, "A2004Analysis.xml"), "data/A2004Analysis.xml")
file.copy(file.path(dataDir, "a2004_O1T2pb05i_A1_A01.fcs"), "data/a2004_O1T2pb05i_A1_A01.fcs")
file.copy(file.path(dataDir, "a2004_O1T2pb05i_A2_A02.fcs"), "data/a2004_O1T2pb05i_A2_A02.fcs")

```{r transformation test}
ws = CytoML::open_flowjo_xml("data/A2004-2026-transformations.gs.wsp")
gs <- CytoML::flowjo_to_gatingset(ws, 
                                  name = "All Samples", 
                                  path = "data",
                                  greedy_match = T)
gs_pop_get_count_fast(gs)
gs_pop_get_gate(gs[[1]],"/asinh-hyperlog")
gh_get_transformations(gs[[1]])

v11_file = "data/2026_02_19.fl11.transform.flowjo"

ws11 <- read_flowjo11_workspace(v11_file)
gs11 <- fj11_to_gatingset(ws11, path = "data",group_name = "Acquired Data", correct_faulty_gate = 256)
gh_get_transformations(gs11[[1]])

gs_save_obj = merge_list_to_gs(gs11)
if(!dir.exists("data/2026_02_19.fl11.transform"))
save_gs(gs = gs_save_obj, path = "data/2026_02_19.fl11.transform")

```


```{r fl11 to gs}

v11_file = "data/2026_02_19.newgates.flowjo"
# v11_file = "data/2026_02_20.Logic.flowjo"
# v11_file = "data/A2004-2026.gsImport.flowjo"

# pretty_print_flowjo(v11_file)

{
  devtools::load_all()

ws11 <- read_flowjo11_workspace(v11_file)
gs11 <- fj11_to_gatingset(ws11, path = "data",group_name = "Acquired Data", correct_faulty_gate = 256)

gs_save_obj = merge_list_to_gs(gs11)
if(!dir.exists("data/2026_02_19.newgates"))
save_gs(gs = gs_save_obj, path = "data/2026_02_19.newgates")

dtList = lapply(gs11,gs_pop_get_count_fast)
dt1 <- do.call(rbind, dtList)
dt1[,c("name","Population", "Count")]
}

gh_pop_get_gate(gs11[[1]], "mDC")
gh_pop_get_gate(gs11[[2]], "MHCII+") %>% str()
gh_get_transformations(gs11[[1]])[["PerCP-CY5-5-A"]]
cf = gh_pop_get_data(gs11[[1]], "single cells", inverse.transform = F)
cf = realize_view(cf)

plot(exprs(cf)[, "PerCP-CY5-5-A"] )


ggcyto::autoplot(gs11[[1]],c("MHCII+"), bins=200, xlim=c(-100, 50000))
ggcyto::autoplot(gs11[[1]],
                 c("/cells/singlets/single cells/MHCII+/Q1-11 - : cd14 +",
  "/cells/singlets/single cells/MHCII+/Q2:APC-A :: CD11c+,PE-CY7-A :: CD14+",
  "/cells/singlets/single cells/MHCII+/Q4:APC-A :: CD11c-,PE-CY7-A :: CD14-",
  "/cells/singlets/single cells/MHCII+/mDC"
  ), bins=200, xlim=c(-100, 50000))

```

```{r fl11 to gs - 2}

v11_file = "data/2026_02_20.Logic.flowjo"
# v11_file = "data/A2004-2026.gsImport.flowjo"

# pretty_print_flowjo(v11_file)

{
  devtools::load_all()

ws11 <- read_flowjo11_workspace(v11_file)
gs11 <- fj11_to_gatingset(ws11, path = "data",group_name = "Acquired Data", correct_faulty_gate = 256)
gs_save_obj = merge_list_to_gs(gs11)
if(!dir.exists("data/2026_02_20.Logic"))
save_gs(gs = gs_save_obj, path = "data/2026_02_20.Logic")

dtList = lapply(gs11,gs_pop_get_count_fast)
dt1 <- do.call(rbind, dtList)
dt1[,c("name","Population", "Count")]
}


ggcyto::autoplot(gs11[[1]],c("/test1"), bins=2000, xlim=c(-100, 50000))
ggcyto::autoplot(gs11[[1]],c("/test2"), bins=2000, xlim=c(-100, 50000))
ggcyto::autoplot(gs11[[1]],c("/test1/r4"), bins=2000, xlim=c(-100, 50000))
ggcyto::autoplot(gs11[[1]],c("/test2/r2"), bins=2000, xlim=c(-100, 50000))
ggcyto::autoplot(gs11[[1]],c("/and gate"), bins=2000, xlim=c(-100, 50000))
ggcyto::autoplot(gs11[[1]],c("/and gate/r2"), bins=2000, xlim=c(-100, 50000))
ggcyto::autoplot(gs11[[1]],c("/or gate"), bins=2000, xlim=c(-100, 50000))
ggcyto::autoplot(gs11[[1]],c("/or gate/r5"), bins=2000, xlim=c(-100, 50000))


```

```{r fl11 to gs}

# v11_file = "data/2026_02_20.Logic.flowjo"
v11_file = "data/A2004-2026.gsImport.flowjo"

# pretty_print_flowjo(v11_file)

{
  devtools::load_all()

ws11 <- read_flowjo11_workspace(v11_file)
gs11 <- fj11_to_gatingset(ws11, path = "data",group_name = "Acquired Data", correct_faulty_gate = 256)
gs_save_obj = merge_list_to_gs(gs11)
if(!dir.exists("data/A2004-2026.gsImport"))
save_gs(gs = gs_save_obj, path = "data/A2004-2026.gsImport")
dtList = lapply(gs11,gs_pop_get_count_fast)
dt1 <- do.call(rbind, dtList)
dt1[,c("name","Population", "Count")]
}

gh_pop_get_gate(gs11[[1]], "mDCs")
gh_pop_get_gate(gs11[[2]], "MHCII+") %>% str()
gh_get_transformations(gs11[[1]])[["PerCP-CY5-5-A"]]
cf = gh_pop_get_data(gs11[[1]], "Single_cells", inverse.transform = F)
cf = realize_view(cf)

tr = gh_get_transformations(gs11[[1]], channel = "PerCP-CY5-5-A")
tr_inv =  gh_get_transformations(gs11[[1]], channel = "PerCP-CY5-5-A",inverse = T)
tr(c(136.300342240011,210.566554318214))
tr_inv(c(136.300342240011,210.566554318214))

tr(exprs(cf)[, "PerCP-CY5-5-A"])
tr_inv(exprs(cf)[, "PerCP-CY5-5-A"])
plot(exprs(cf)[, "PerCP-CY5-5-A"] )
tr(105)


ggcyto::autoplot(gs11[[1]],c("MHCII+"), bins=200, xlim=c(-100, 50000))
ggcyto::autoplot(gs11[[1]],
                 c("/cells/Singlets/Single_cells/MHCII+/Q5: CD11c- , CD14+",
  "/cells/Singlets/Single_cells/MHCII+/Q6: CD11c+ , CD14+",
  "/cells/Singlets/Single_cells/MHCII+/mDCs",
  "/cells/Singlets/Single_cells/MHCII+/Q8: CD11c- , CD14-"
  ), bins=200, xlim=c(-100, 50000))

```

```{r}
test_name <- tools::file_path_sans_ext(basename(v11_file))
result <- list(
  test_name = test_name,
  success = FALSE,
  error = NULL,
  stats_before = NULL,
  stats_after = NULL,
  gate_count = 0,
  sample_count = 0,
  exported_file = NULL
)




ws = CytoML::open_flowjo_xml("data/A2004-2026.wsp")
v11_file = "data/A2004-2026.flowjo"
ws11 <- read_flowjo11_workspace(v11_file)
gs <- CytoML::flowjo_to_gatingset(ws, 
                                  name = "All Samples", 
                                  path = "data",
                                  greedy_match = T)
gh_get_pop_paths(gs, path = "auto")
  pretty_print_flowjo(v11_file)


gs11 <- fj11_to_gatingset(ws11, path = "data",group_name = "Acquired Data", correct_faulty_gate = 256)
lapply(gs11,flowWorkspace::recompute)
gs_save_obj = merge_list_to_gs(gs11)
if(!dir.exists("data/A2004-2026_gs"))
save_gs(gs = gs_save_obj, path = "data/A2004-2026_gs")


# Select columns before merge
dt1 <- lapply(gs11, flowWorkspace::gs_pop_get_count_fast) %>%  data.table::rbindlist()  %>%
  .[, .(name, Population, Count, ParentCount)]

dt2 <- flowWorkspace::gs_pop_get_count_fast(gs)[, .(name, Population, Count, ParentCount)]

# Clean names
dt1[, base_name := sub("_[0-9]+$", "", name)]
dt2[, base_name := sub("_[0-9]+$", "", name)]

# Merge
combined <- merge(dt1, dt2,
                  by = c("base_name", "Population"),
                  suffixes = c("_gs11", "_gs"),all = T)
combined[, `:=`(
  Ratio_gs11 = Count_gs11 / ParentCount_gs11,
  Ratio_gs = Count_gs / ParentCount_gs
)]

# Or if you want to merge with your existing combined table:
combined_gs = combined %>% 
  select(c("base_name", "Population", "Ratio_gs", "Ratio_gs11",  "Count_gs", "Count_gs11")) %>%
  rename(
    gs10_freq = Ratio_gs,
    gs11_freq = Ratio_gs11,
    gs10_cnt  = Count_gs,
    gs11_cnt = Count_gs11
  )



```


```{r flowjoStats}

Statistics_fj11 <- read_csv("data/Statistics.fj11.csv")
statsFL10 <- read_csv("data/statsFL10.csv")


# Clean Statistics_fj11 (remove summary rows and reshape)
clean_fj11 <- Statistics_fj11 %>%
  filter(!`File Name` %in% c("Mean", "Median", "Max", "SD", "Min", "CV")) %>%
  select(`File Name`, matches("Freq\\. of Parent|Count")) %>%
  pivot_longer(
    cols = -`File Name`,
    names_to = "measure",
    values_to = "value"
  ) %>%
  mutate(
    # Extract population path
    Population = str_replace(measure, "Ungated", "/") %>%
      str_replace(" Freq\\. of Parent.*", "") %>%
      str_replace(" Count.*", "") %>%
      str_replace_all(" / ", "/") %>%
      str_replace_all("//", "/") %>%
      str_trim(),
    # Determine if it's frequency or count
    metric_type = ifelse(str_detect(measure, "Freq"), "Frequency", "Count")
  ) %>%
  select(File_Name = `File Name`, Population, metric_type, value) %>%
  pivot_wider(
    names_from = metric_type,
    values_from = value,
    names_prefix = "fj11_"
  ) %>%
  rename(base_name = File_Name)

# Clean statsFj10 (already in better format, but needs reshaping)
clean_fj10 <- statsFj10 %>%
  rename(base_name = ...1, "/_Count" = Count) %>%  # FIX HERE
  select(-matches("^\\.\\.\\.")) %>%  # Remove any other ...XX columns
  pivot_longer(
    cols = -base_name,
    names_to = "measure",
    values_to = "value"
  ) %>%
  mutate(
    # Extract population path
    Population = str_extract(measure, "^[^|]+") %>% str_trim(),
    Population = paste0("/", str_replace_all(Population, "/", "/")) %>%
      str_replace("/_Count", ""),
    # Determine metric type
    metric_type = case_when(
      str_detect(measure, "Freq\\.") ~ "Frequency",
      str_detect(measure, "Count") ~ "Count",
      TRUE ~ "Other"
    )
  ) %>%
  filter(metric_type %in% c("Frequency", "Count")) %>%
  select(base_name, Population, metric_type, value) %>%
  pivot_wider(
    names_from = metric_type,
    values_from = value,
    names_prefix = "fj10_"
  ) %>%
  mutate(
    fj10_Frequency = fj10_Frequency / 100  # Convert percentage to proportion
  )

# Join everything together
combined_all <- clean_fj11 %>%
  full_join(clean_fj10, by = c("base_name", "Population")) %>%
  arrange(base_name, Population) %>% 
  rename(
    fj11_cnt  = fj11_Count,
    fj11_freq = fj11_Frequency,
    fj10_cnt  = fj10_Count,
    fj10_freq = fj10_Frequency
  )

# View the result
head(combined_all)


```

```{r}

combined_final <- combined_gs %>%
  left_join(
    combined_all,
    by = c("base_name", "Population")
  )

combined_final %>% head()


```

```{r correlation matrix}

# Create correlation matrix
freq_data <- combined_final %>%
  filter(!is.na(gs10_freq), !is.na(gs11_freq), !is.na(fj10_freq), !is.na(fj11_freq)) %>%
  select(gs10_freq, gs11_freq, fj10_freq, fj11_freq)

# Correlation pairs plot
p1 <- ggpairs(freq_data,
              columnLabels = c("GS v10", "GS v11", "FJ v10", "FJ v11"),
              upper = list(continuous = wrap("cor", size = 5, color = "black")),
              lower = list(continuous = wrap("points", alpha = 0.3, size = 0.5)),
              diag = list(continuous = wrap("densityDiag", alpha = 0.5))) +
  theme_bw() +
  labs(title = "Method Comparison: Correlation Matrix")

print(p1)
```

```{r bland altman pairwise}


# Calculate all pairwise differences
ba_data <- combined_final %>%
  filter(!is.na(gs10_freq), !is.na(gs11_freq), !is.na(fj10_freq), !is.na(fj11_freq)) %>%
  mutate(
    # GS10 vs GS11
    mean_gs = (gs10_freq + gs11_freq) / 2,
    diff_gs = gs11_freq - gs10_freq,
    
    # FJ10 vs FJ11
    mean_fj = (fj10_freq + fj11_freq) / 2,
    diff_fj = fj11_freq - fj10_freq,
    
    # GS10 vs FJ10
    mean_v10 = (gs10_freq + fj10_freq) / 2,
    diff_v10 = fj10_freq - gs10_freq,
    
    # GS11 vs FJ11
    mean_v11 = (gs11_freq + fj11_freq) / 2,
    diff_v11 = fj11_freq - gs11_freq
  )

# Function to create BA plot
create_ba_plot <- function(data, mean_col, diff_col, title) {
  mean_diff <- mean(data[[diff_col]], na.rm = TRUE)
  sd_diff <- sd(data[[diff_col]], na.rm = TRUE)
  
  ggplot(data, aes(x = .data[[mean_col]], y = .data[[diff_col]])) +
    geom_point(alpha = 0.4, size = 2) +
    geom_hline(yintercept = 0, color = "gray50", linetype = "dashed") +
    geom_hline(yintercept = mean_diff, color = "blue", size = 1) +
    geom_hline(yintercept = mean_diff + 1.96 * sd_diff, 
               color = "red", linetype = "dashed") +
    geom_hline(yintercept = mean_diff - 1.96 * sd_diff, 
               color = "red", linetype = "dashed") +
    annotate("text", x = max(data[[mean_col]], na.rm = TRUE) * 0.8, 
             y = mean_diff, label = sprintf("Bias: %.4f", mean_diff), 
             vjust = -0.5, color = "blue") +
    annotate("text", x = max(data[[mean_col]], na.rm = TRUE) * 0.8, 
             y = mean_diff + 1.96 * sd_diff, 
             label = sprintf("+1.96 SD: %.4f", mean_diff + 1.96 * sd_diff), 
             vjust = -0.5, color = "red", size = 3) +
    labs(title = title,
         x = "Mean of Methods",
         y = "Difference") +
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, size = 10))
}

# Create all BA plots
p_gs <- create_ba_plot(ba_data, "mean_gs", "diff_gs", "GS: v11 - v10")
p_fj <- create_ba_plot(ba_data, "mean_fj", "diff_fj", "FJ: v11 - v10")
p_v10 <- create_ba_plot(ba_data, "mean_v10", "diff_v10", "v10: FJ - GS")
p_v11 <- create_ba_plot(ba_data, "mean_v11", "diff_v11", "v11: FJ - GS")

# Combine into one figure
(p_gs + p_fj) / (p_v10 + p_v11) +
  plot_annotation(title = "Bland-Altman Analysis: Method Comparisons",
                  theme = theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")))

```


3. Summary Statistics Table (Main Table)

```{r}

# Calculate comprehensive statistics
comparison_summary <- tibble(
  Comparison = c("GS v10 vs v11", "FJ v10 vs v11", "v10: GS vs FJ", "v11: GS vs FJ"),
  
  # Sample size
  N = c(
    sum(!is.na(ba_data$diff_gs)),
    sum(!is.na(ba_data$diff_fj)),
    sum(!is.na(ba_data$diff_v10)),
    sum(!is.na(ba_data$diff_v11))
  ),
  
  # Correlation
  `Pearson r` = c(
    cor(ba_data$gs10_freq, ba_data$gs11_freq, use = "complete.obs"),
    cor(ba_data$fj10_freq, ba_data$fj11_freq, use = "complete.obs"),
    cor(ba_data$gs10_freq, ba_data$fj10_freq, use = "complete.obs"),
    cor(ba_data$gs11_freq, ba_data$fj11_freq, use = "complete.obs")
  ),
  
  # Bias
  `Mean Bias` = c(
    mean(ba_data$diff_gs, na.rm = TRUE),
    mean(ba_data$diff_fj, na.rm = TRUE),
    mean(ba_data$diff_v10, na.rm = TRUE),
    mean(ba_data$diff_v11, na.rm = TRUE)
  ),
  
  # Precision
  `SD` = c(
    sd(ba_data$diff_gs, na.rm = TRUE),
    sd(ba_data$diff_fj, na.rm = TRUE),
    sd(ba_data$diff_v10, na.rm = TRUE),
    sd(ba_data$diff_v11, na.rm = TRUE)
  ),
  
  # Limits of agreement
  `95% LoA Lower` = `Mean Bias` - 1.96 * SD,
  `95% LoA Upper` = `Mean Bias` + 1.96 * SD,
  
  # Agreement within thresholds
  `Within ±5%` = c(
    mean(abs(ba_data$diff_gs) <= 0.05, na.rm = TRUE) * 100,
    mean(abs(ba_data$diff_fj) <= 0.05, na.rm = TRUE) * 100,
    mean(abs(ba_data$diff_v10) <= 0.05, na.rm = TRUE) * 100,
    mean(abs(ba_data$diff_v11) <= 0.05, na.rm = TRUE) * 100
  )
) %>%
  mutate(across(where(is.numeric), ~round(., 4)))

# Publication-ready table
comparison_summary %>%
  kable(format = "html", 
        caption = "Table 1. Method Comparison Statistics",
        digits = 4) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 3, "Bland-Altman Analysis" = 4, "Agreement" = 1))

# For LaTeX/PDF:
comparison_summary %>%
  kable(format = "latex", booktabs = TRUE, 
        caption = "Method Comparison Statistics") %>%
  kable_styling(latex_options = c("striped", "hold_position"))
```

```{r}
combined_all <- clean_fj11 %>%
  full_join(clean_fj10, by = c("base_name", "Population")) %>%
  arrange(base_name, Population) %>%
  mutate(
    # Average of two methods
    mean_frequency = (fj11_Frequency + fj10_Frequency) / 2,
    
    # Difference between methods
    diff_frequency = fj11_Frequency - fj10_Frequency,
    
    # Percent difference relative to mean
    percent_diff = (diff_frequency / mean_frequency) * 100,
    
    # Agreement within ±5% or ±10%
    within_5pct = abs(diff_frequency) <= 0.05,
    within_10pct = abs(diff_frequency) <= 0.10
  )

# Bland-Altman plot

ggplot(combined_all, aes(x = mean_frequency, y = diff_frequency)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, color = "blue", linetype = "dashed") +
  geom_hline(yintercept = mean(combined_all$diff_frequency, na.rm = TRUE), 
             color = "red") +
  geom_hline(yintercept = mean(combined_all$diff_frequency, na.rm = TRUE) + 
             1.96 * sd(combined_all$diff_frequency, na.rm = TRUE),
             color = "red", linetype = "dashed") +
  geom_hline(yintercept = mean(combined_all$diff_frequency, na.rm = TRUE) - 
             1.96 * sd(combined_all$diff_frequency, na.rm = TRUE),
             color = "red", linetype = "dashed") +
  labs(title = "Bland-Altman Plot: Method Comparison",
       x = "Mean of fj10 and fj11",
       y = "Difference (fj11 - fj10)") +
  theme_bw()
```



```{r, eval=T}

gh_get_pop_paths(gs11[[1]], path = "auto")

ggcyto::autoplot(gs,"/Lymphocytes-test", bins=2000)
ggcyto::autoplot(gs11[[1]],"/Lymphocytes-test", bins=2000)
ggcyto::autoplot(gs11[[1]],"/cells/Singlets/Single_cells/MHCII+", bins=2000)
ggcyto::autoplot(gs11[[1]],"Q8: CD11c- , CD14-", bins=2000)

gh_pop_get_gate(gs11[[1]], "MHCII+")
ggcyto::autoplot(gs11[[1]],c("MHCII+"), bins=2000)

ggcyto::autoplot(gs11[[1]],c("mDCs", "Q5: CD11c- , CD14+"), bins=2000)


plot_list <- lapply(gh_get_pop_paths(gs11[[1]], path = "auto")[-1], function(pop) {
  message(pop)
  # Get the gate for this population
  gate <- gh_pop_get_gate(gs11[[1]], pop)
  
  # Get parent data (since gate is defined on parent)
  parent_pop <- dirname(pop)  # Get parent path
  if (parent_pop == ".") parent_pop <- "root"
  
  parent_data <- exprs(gh_pop_get_data(gs11[[1]], parent_pop))
  
  # Extract channel names from gate
  if (inherits(gate, "rectangleGate")) {
    channels <- parameters(gate) %>% names()
  } else if (inherits(gate, "polygonGate") || inherits(gate, "ellipsoidGate")) {
    channels <- gate@parameters %>% unlist() %>% names()
  } else {
    channels <- NULL
  }
  if(length(channels) == 1){
    channels = c(channels, "SSC-A")
  }
  # Create base plot
  p <- ggcyto::autoplot(gs11[[1]], pop, bins = 2000) + 
    ggtitle(pop)
  
  # Add dynamic limits only for 2D gates
  if (!is.null(channels) && length(channels) == 2) {
    tryCatch({
      p <- p +
        coord_cartesian(xlim = c(0, quantile(parent_data[, channels[1]], 0.99, na.rm = TRUE)),
        ylim = c(0, quantile(parent_data[, channels[2]], 0.99, na.rm = TRUE)))
    }, error = function(e) {
      message("Could not set limits for ", pop, ": ", e$message)
    })
  }
  
  return(p)
})
plot_list
# Print all
# gh_pop_get_gate(gs11[[1]],pop)

lapply(gh_get_pop_paths(gs11[[1]], path = "auto")[-1], function(pop) {
  message(str(gh_pop_get_gate(gs11[[1]],pop)))
})


flowWorkspace::recompute(gs)
gs_pop_get_count_fast(gs)
gs_pop_get_count_fast(gs11[[1]])
gate11 = gs_pop_get_gate(gs11, "cells")
gate = gs_pop_get_gate(gs, "cells")
ggcyto::autoplot(gs11[[1]],"/cells/Singlets", bins=2000)


samples <- ws11$samples
result$sample_count <- length(samples)
cat("  - Found", result$sample_count, "sample(s)\n")

markernames(gs)
```





```{r}



# Read the file - readr handles the complex headers better than base R
statsFj11 <- read_csv("data/Statistics.fj11.csv", 
               name_repair = "unique",  # Ensures unique column names
               na = c("", "NA", "−1.00"))  # Handle the unicode minus in missing values
statsFj11 = statsFj11[c(1,2),]

# Read the file - readr handles the complex headers better than base R
statsFj10 <- read_csv("data/statsFL10.csv", 
               name_repair = "unique",  # Ensures unique column names
               na = c("", "NA", "−1.00"))  # Handle the unicode minus in missing values
statsFj10 = statsFj10[c(1,2),]

colnames(statsFj10)
colnames(statsFj11)


# Apply to both datasets
new_names_fj10 <- standardize_column_names(colnames(statsFj10))
new_names_fj11 <- standardize_column_names(colnames(statsFj11))

# Rename the dataframes
statsFj10_renamed <- statsFj10
colnames(statsFj10_renamed) <- new_names_fj10

statsFj11_renamed <- statsFj11
colnames(statsFj11_renamed) <- new_names_fj11

# Check which columns are now common between both
common_cols <- intersect(new_names_fj10, new_names_fj11)
only_fj10 <- setdiff(new_names_fj10, new_names_fj11)
only_fj11 <- setdiff(new_names_fj11, new_names_fj10)

# Summary report
cat("Column Renaming Summary\n")
cat("======================\n")
cat("Common columns:", length(common_cols), "\n")
cat("Only in Fj10:", length(only_fj10), "\n")
cat("Only in Fj11:", length(only_fj11), "\n\n")

cat("Common columns:\n")
print(common_cols)

cat("\n\nOnly in Fj10:\n")
print(only_fj10)

cat("\n\nOnly in Fj11:\n")
print(only_fj11)

# View the renamed column names side by side
comparison <- data.frame(
  original_fj10 = colnames(statsFj10),
  new_fj10 = new_names_fj10,
  original_fj11 = c(colnames(statsFj11), rep(NA, length(new_names_fj10) - length(colnames(statsFj11)))),
  new_fj11 = c(new_names_fj11, rep(NA, length(new_names_fj10) - length(new_names_fj11)))
)

print(comparison)
```


**For each test case, record:**
1. Gate coordinates (screenshot or manual recording)
2. Population counts and percentages
3. Transformation parameters used
4. Compensation matrix (if applicable)

========

### Phase 2: Export via CyFj11 Workflow

**I will provide R code for:**

```{r}
gs_dir <- "data/2026_02_19.fl11.transform"
gs_dir <- "data/2026_02_19.newgates"
gs_dir <- "data/2026_02_20.Logic"
gs_dir <- "data/A2004-2026.gsImport"
gs_dir <- "data/A2004-2026_gs"

gs11 <- load_gs(gs_dir)
# Step 2: Export to v10

export_flowjo10_workspace(
gating_set = gs11,
output_path = paste0(gs_dir, "_exported.wsp")
)

# Step 3: Extract statistics for comparison
stats <- getPopStats(gs)
write.csv(stats, "test_case_X_r_stats.csv")
```



**Output files needed:**
- `test_case_X_exported.wsp` (v10 workspace)
- `test_case_X_r_stats.csv` (population statistics from R)

========================================

### Phase 3: Import Back to FlowJo v11

**You will:**
1. Open FlowJo v11 (fresh instance)
2. Import `test_case_X_exported.wsp`
3. Record what happens:
- Does it open without errors?
- Are all gates visible?
- Are gate coordinates preserved?
- Do population percentages match?

**Data to collect for each test case:**

| Metric | How to Measure | Tolerance |
|--------|----------------|-----------|
| Gate coordinates | Compare FlowJo display (screenshot) | ±0.1% of axis range |
| Population counts | FlowJo Statistics view vs R stats | ±0.5% relative difference |
| Population percentages | FlowJo vs R | ±0.5% absolute difference |
| Gate hierarchy | Visual inspection of population tree | Must match exactly |
| Boolean logic | Verify AND/OR/NOT relationships | 100% cell membership agreement |

========================================

### Phase 4: Roundtrip Validation (Optional but Powerful)

**If you have time, also test:**

1. **Full roundtrip:** v11 → R → v10 → FlowJo v11 → re-export as v11 → compare to original
2. **Feature degradation check:** What v11 features are lost?
- Layout annotations?
- Statistical overlays?
- Batch processing metadata?
- Derived parameters with complex formulas?

========================================

## Pass/Fail Criteria

### Critical Success (Must Pass for ≥8/10 test cases):
- ✓ Workspace opens in FlowJo v11 without errors
- ✓ All gates display correctly
- ✓ Population counts within ±0.5%

### Full Success (Reviewer 2 Standard):
- ✓ Gate coordinates within ±0.1% of axis range
- ✓ Boolean gate logic 100% preserved
- ✓ Compensation matrices mathematically identical
- ✓ Transformation parameters yield equivalent population distributions (KS test p>0.05)

========================================

## Deliverables for Manuscript

1. **Summary Table:**

| Test Case | n Samples | n Gates | Gate Types | Opened OK? | Count Match? | Notes |
|-----------|-----------|---------|------------|------------|--------------|-------|
| TC-1 | 1 | 3 | Rectangle | ✓ | ✓ | |
| TC-2 | 1 | 3 | Polygon | ? | ? | |
| ... | | | | | | |

2. **Quantitative Results:**
- Mean absolute error in gate coordinates
- Concordance correlation coefficient for population frequencies
- Percentage of gates successfully converted by type

3. **Failure Documentation:**
- Which gate types (if any) fail?
- What error messages appear?
- What features are lost?

========================================

## Required Data Files

I can generate synthetic test data or you can provide real (de-identified) FCS files. What's your preference?

**Option A:** I create mock FCS files with known population structures (guaranteed cell counts in each gate)
**Option B:** You provide real FCS files from your lab
**Option C:** Use public FlowRepository datasets

========================================

## Timeline Estimate

| Phase | Time Required | Who |
|-------|---------------|-----|
| Create test workspaces | 2-4 hours | You (FlowJo) |
| Run CyFj11 export | 30 min | Me (R code) |
| Validate in FlowJo v11 | 2-3 hours | You (FlowJo) |
| Compile results | 1 hour | Together |
| **Total** | **~1 day** | **Split effort** |

========================================

## Next Steps

1. **Confirm you have FlowJo v11 access** and time for ~1 day of testing
2. **Choose data source** (mock/real/public)
3. **Decide scope** (minimum 5 test cases vs all 10)
4. I'll prepare the complete R scripts for Phase 2

========================================

## Alternative if FlowJo v11 Access is Limited

If you cannot run FlowJo, options are:

1. **Collaborator approach:** Find a FlowJo user willing to run the test protocol
2. **Literature-based approach:** Cite BD Biosciences documentation that v11 opens v10 (but reviewer may reject this)
3. **Honest limitation statement:** State that compatibility testing is ongoing and planned for future validation (weakens manuscript significantly)

Reviewer 2's position: "Your workflow relies on the critical assumption that FlowJo v11 can import v10 workspace files without data loss, yet you provide neither empirical verification nor citations..."

**Empirical verification is strongly preferred.**

========================================

*Protocol version 1.0 - 2026-02-04*
