---
title: 'Developer Test Suite: export_flowjo10_workspace'
author: "Incremental Testing with FlowJo Validation"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: false
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  fig.width = 10,
  fig.height = 7
)
devtools::load_all()
```

# Overview

Test key functionality 

---

# Configuration

**IMPORTANT:** Set your local test directory here!

```{r config}
# Set this to where you want test files saved
# All tests will create subdirectories under this path
BASE_TEST_DIR <- "/pasteur/helix/projects/scBiomarkers/bernd/cytometry/CyFl11/flowjo_export_tests"  # CHANGE THIS TO YOUR PREFERRED LOCATION

# Alternative: use current working directory
# BASE_TEST_DIR <- file.path(getwd(), "test_output")

# Create base directory
dir.create(BASE_TEST_DIR, showWarnings = FALSE, recursive = TRUE)

cat(paste(rep("=", 70), collapse = ""), "\n")
cat("TEST CONFIGURATION\n")
cat(paste(rep("=", 70), collapse = ""), "\n")
cat("Base test directory:", BASE_TEST_DIR, "\n")
cat("Absolute path:", normalizePath(BASE_TEST_DIR, mustWork = FALSE), "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

# Create README for FlowJo testing
readme_path <- file.path(BASE_TEST_DIR, "README.txt")
writeLines(c(
  "FlowJo Export Test Files",
  "=======================",
  "",
  paste("Generated:", Sys.time()),
  "",
  "Directory Structure:",
  "  test01/ - Minimal export (no gates, linear only)",
  "  test02/ - Single 1D rectangle gate (linear)",
  "  test03/ - Single 2D rectangle gate (linear)",
  "  test04/ - Polygon gate (linear)",
  "  test05/ - Ellipsoid gate (linear)",
  "  test06/ - Biexponential transformation",
  "  test07/ - Log transformation",
  "  test08/ - Arcsinh transformation",
  "  test09/ - Hierarchical gates (2 levels)",
  "  test10/ - Boolean AND gate",
  "",
  "How to Test in FlowJo:",
  "  1. Open FlowJo v10",
  "  2. File → Import → Workspace",
  "  3. Select export.wsp from any test directory",
  "  4. Verify gates appear correctly",
  "  5. Check population statistics match expected values",
  "",
  "Each test directory contains:",
  "  - *.fcs - Original FCS files",
  "  - export.wsp - FlowJo v10 workspace",
  "  - test_info.txt - Details about what was tested"
), readme_path)

cat("✓ Created README.txt in test directory\n\n")
```

---

# Setup

```{r libraries, message=FALSE, warning=FALSE}
library(flowCore)
library(flowWorkspace)
library(CytoML)
library(openCyto)
library(Biobase)
library(ggcyto)
library(gridExtra)

# Source your export function
# source("R/export_flowjo10_workspace.R")
```

## Utility Functions

```{r utilities}
#' Create minimal synthetic FCS
create_test_fcs <- function(n = 10000, seed = 123, n_channels = 4) {
  set.seed(seed)
  
  # Base scatter channels
  fsc_a <- rnorm(n, mean = 120000, sd = 30000)
  fsc_h <- fsc_a * runif(n, 0.9, 1.1)
  ssc_a <- rnorm(n, mean = 80000, sd = 20000)
  ssc_h <- ssc_a * runif(n, 0.9, 1.1)
  
  # Fluorescence channels (bimodal)
  fitc_a <- c(rlnorm(n * 0.7, 2, 0.8), rlnorm(n * 0.3, 8, 0.5))[1:n]
  pe_a <- c(rlnorm(n * 0.6, 2.5, 0.7), rlnorm(n * 0.4, 7.5, 0.6))[1:n]
  apc_a <- c(rlnorm(n * 0.8, 1.8, 0.9), rlnorm(n * 0.2, 7.2, 0.5))[1:n]
  percpcy55_a <- c(rlnorm(n * 0.75, 2.2, 0.75), rlnorm(n * 0.25, 7.8, 0.55))[1:n]
  
  clamp <- function(x) pmin(pmax(x, 0), 262144)
  
  mat <- cbind(
    clamp(fsc_a), clamp(fsc_h), clamp(ssc_a), clamp(ssc_h),
    clamp(fitc_a), clamp(pe_a), clamp(apc_a), clamp(percpcy55_a)
  )
  colnames(mat) <- c("FSC-A", "FSC-H", "SSC-A", "SSC-H", 
                     "FITC-A", "PE-A", "APC-A", "PerCP-Cy5-5-A")
  
  params <- AnnotatedDataFrame(data = data.frame(
    name = colnames(mat),
    desc = c("FSC-A", "FSC-H", "SSC-A", "SSC-H", "CD3", "CD4", "CD8", "CD14"),
    range = rep(262144, ncol(mat)),
    minRange = rep(0, ncol(mat)),
    maxRange = rep(262144, ncol(mat)),
    row.names = colnames(mat),
    stringsAsFactors = FALSE
  ))
  
  ff <- new("flowFrame",
            exprs = mat,
            parameters = params,
            description = list(
              `$FIL` = sprintf("test_sample_%03d.fcs", seed),
              FILENAME = sprintf("test_sample_%03d.fcs", seed),
              `$TOT` = as.character(n),
              `$PAR` = as.character(ncol(mat))
            ))
  return(ff)
}

#' Save test info to file
save_test_info <- function(test_dir, test_name, description, gate_info = NULL, 
                          transform_info = NULL, expected_counts = NULL) {
  info_path <- file.path(test_dir, "test_info.txt")
  
  lines <- c(
    paste("Test:", test_name),
    paste("Date:", Sys.time()),
    "",
    "DESCRIPTION:",
    description,
    ""
  )
  
  if (!is.null(gate_info)) {
    lines <- c(lines, "GATES:", gate_info, "")
  }
  
  if (!is.null(transform_info)) {
    lines <- c(lines, "TRANSFORMATIONS:", transform_info, "")
  }
  
  if (!is.null(expected_counts)) {
    lines <- c(lines, "EXPECTED COUNTS:", expected_counts, "")
  }
  
  lines <- c(lines,
    "",
    "FILES:",
    "  export.wsp - FlowJo v10 workspace (OPEN THIS IN FLOWJO)",
    "  *.fcs - FCS data files",
    "",
    "FLOWJO VALIDATION STEPS:",
    "  1. Open FlowJo v10",
    "  2. File → Import → Workspace",
    "  3. Select 'export.wsp' from this directory",
    "  4. Verify gates display correctly",
    "  5. Check population counts match expected values above",
    "  6. Verify transformations are applied correctly"
  )
  
  writeLines(lines, info_path)
}

#' Inspect GatingSet structure
inspect_gs <- function(gs) {
  cat(paste(rep("=", 70), collapse = ""), "\n")
  cat("GATINGSET INSPECTION\n")
  cat(paste(rep("=", 70), collapse = ""), "\n\n")
  
  cat("Samples:", paste(sampleNames(gs), collapse = ", "), "\n")
  cat("Populations:", length(gs_get_pop_paths(gs)), "\n\n")
  
  # Population hierarchy
  cat("Population Hierarchy:\n")
  pops <- gs_get_pop_paths(gs)
  stats <- gs_pop_get_count_fast(gs)
  
  for (pop in pops) {
    depth <- length(strsplit(pop, "/")[[1]]) - 1
    indent <- paste(rep("  ", depth), collapse = "")
    
    # Get count
    pop_stats <- stats[stats$Population == pop, ]
    if (nrow(pop_stats) > 0) {
      count <- sum(pop_stats$Count)
    } else {
      count <- 0
    }
    
    # Get gate type
    if (pop != "root") {
      gate <- gs_pop_get_gate(gs, pop)[[1]]
      gate_type <- class(gate)[1]
      
      # Get parameters
      params <- tryCatch({
        p <- parameters(gate)
        if (is.character(p)) p else names(p)
      }, error = function(e) character(0))
      
      cat(sprintf("%s├─ %s [%s on %s] (n=%d)\n", 
                  indent, 
                  basename(pop), 
                  gate_type,
                  paste(params, collapse = ", "),
                  count))
    } else {
      cat(sprintf("%s%s (n=%d)\n", indent, pop, count))
    }
  }
  
  # Transformations
  cat("\nTransformations:\n")
  gh <- gs[[1]]
  trans <- gh_get_transformations(gh)
  
  if (length(trans) > 0) {
    for (ch in names(trans)) {
      trans_type <- attr(trans[[ch]], "type")
      if (is.null(trans_type)) trans_type <- "unknown"
      cat(sprintf("  %s: %s\n", ch, trans_type))
    }
  } else {
    cat("  None (all linear)\n")
  }
  
  cat("\n")
  
  # Return stats for saving
  return(stats)
}

#' Plot GatingSet
plot_gs <- function(gs, max_pops = 6) {
  pops <- gs_get_pop_paths(gs)
  pops <- pops[pops != "root"]
  
  if (length(pops) == 0) {
    cat("No gates to plot\n")
    return(NULL)
  }
  
  # Limit number of plots
  pops <- head(pops, max_pops)
  
  plot_list <- list()
  
  for (pop in pops) {
    tryCatch({
      gate <- gs_pop_get_gate(gs, pop)[[1]]
      
      # Get parameters
      params <- tryCatch({
        p <- parameters(gate)
        if (is.character(p)) p else names(p)
      }, error = function(e) character(0))
      
      if (length(params) >= 2) {
        # 2D plot
        p <- ggcyto(gs, aes(x = !!sym(params[1]), y = !!sym(params[2])), 
                    subset = "root") +
          geom_hex(bins = 64) +
          geom_gate(pop) +
          ggtitle(pop) +
          theme_minimal()
        plot_list[[pop]] <- p
      } else if (length(params) == 1) {
        # 1D plot
        p <- ggcyto(gs, aes(x = !!sym(params[1])), subset = "root") +
          geom_density(fill = "lightblue") +
          geom_gate(pop) +
          ggtitle(pop) +
          theme_minimal()
        plot_list[[pop]] <- p
      }
    }, error = function(e) {
      cat("Could not plot", pop, ":", e$message, "\n")
    })
  }
  
  if (length(plot_list) > 0) {
    do.call(grid.arrange, c(plot_list, ncol = 2))
  }
}

#' Show XML snippet
show_xml_snippet <- function(xml_path, pattern = NULL, context = 5) {
  if (!file.exists(xml_path)) {
    cat("XML file not found\n")
    return(invisible())
  }
  
  lines <- readLines(xml_path)
  
  if (is.null(pattern)) {
    # Show first N lines
    cat("First 30 lines of XML:\n")
    cat(paste(rep("-", 70), collapse = ""), "\n")
    cat(paste(head(lines, 30), collapse = "\n"), "\n")
    cat(paste(rep("-", 70), collapse = ""), "\n")
  } else {
    # Show lines matching pattern with context
    matches <- grep(pattern, lines)
    if (length(matches) > 0) {
      cat(sprintf("Lines matching '%s':\n", pattern))
      cat(paste(rep("-", 70), collapse = ""), "\n")
      
      for (m in head(matches, 5)) {
        start <- max(1, m - context)
        end <- min(length(lines), m + context)
        
        cat(sprintf("\n--- Around line %d ---\n", m))
        cat(paste(lines[start:end], collapse = "\n"), "\n")
      }
      cat(paste(rep("-", 70), collapse = ""), "\n")
    } else {
      cat(sprintf("No lines matching '%s'\n", pattern))
    }
  }
}

#' Test roundtrip
test_roundtrip <- function(gs_orig, export_path, fcs_dir, tolerance = 1.0) {
  cat("\n")
  cat(paste(rep("=", 70), collapse = ""), "\n")
  cat("ROUNDTRIP VALIDATION\n")
  cat(paste(rep("=", 70), collapse = ""), "\n\n")
  
  tryCatch({
    # Read back
    ws <- CytoML::open_flowjo_xml(export_path)
    cat("✓ Workspace opened\n")
    
    gs_back <- CytoML::flowjo_to_gatingset(ws, name = "All Samples", 
                                            path = fcs_dir)
    cat("✓ GatingSet reconstructed\n")
    
    # Compare
    stats_orig <- gs_pop_get_count_fast(gs_orig)
    stats_back <- gs_pop_get_count_fast(gs_back)
    
    comparison <- merge(
      stats_orig[, c("name", "Population", "Count")],
      stats_back[, c("name", "Population", "Count")],
      by = c("name", "Population"),
      suffixes = c("_orig", "_back"),
      all = TRUE
    )
    
    comparison$diff <- comparison$Count_back - comparison$Count_orig
    comparison$pct_diff <- ifelse(
      comparison$Count_orig > 0,
      abs(comparison$diff) / comparison$Count_orig * 100,
      0
    )
    
    max_diff <- max(comparison$pct_diff, na.rm = TRUE)
    
    print(knitr::kable(comparison, digits = 2))
    
    cat(sprintf("\nMax difference: %.2f%%\n", max_diff))
    cat(sprintf("Tolerance: %.2f%%\n", tolerance))
    
    if (max_diff <= tolerance) {
      cat("\n✓✓✓ ROUNDTRIP PASSED ✓✓✓\n")
      return(TRUE)
    } else {
      cat("\n✗✗✗ ROUNDTRIP FAILED ✗✗✗\n")
      cat("\nPopulations exceeding tolerance:\n")
      print(comparison[comparison$pct_diff > tolerance, ])
      return(FALSE)
    }
  }, error = function(e) {
    cat("✗ Roundtrip error:", e$message, "\n")
    return(FALSE)
  })
}
```

---

# TEST 1: Minimal - No Gates, Linear Only

**Goal:** Export empty GatingSet (root only, no transformations)

## 1.1 Setup Test Data

```{r test01_setup}
test_name <- "TEST 1: Minimal (no gates, linear)"
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat(test_name, "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

# Create test directory
test_dir <- file.path(BASE_TEST_DIR, "test01")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

# Create test data
ff <- create_test_fcs(n = 5000, seed = 1)

fcs_path <- file.path(test_dir, "sample01.fcs")
write.FCS(ff, fcs_path)

# Create GatingSet
fs <- read.flowSet(fcs_path)
gs <- GatingSet(fs)

cat("✓ Test data created\n")
cat("  Directory:", test_dir, "\n")
cat("  FCS:", fcs_path, "\n")
cat("  Events:", nrow(ff), "\n\n")

# Save test info
save_test_info(
  test_dir, 
  test_name,
  description = "Minimal test with no gates and no transformations. Only root population.",
  gate_info = "None - only root population",
  transform_info = "None - all channels linear",
  expected_counts = sprintf("root: %d events", nrow(ff))
)

cat("✓ Test info saved to:", file.path(test_dir, "test_info.txt"), "\n")
```

## 1.4 Export

```{r test01_export}
export_path <- file.path(test_dir, "export.wsp")

cat("Exporting to:", export_path, "\n\n")
success <- export_flowjo10_workspace(gs, export_path)


```

## 1.6 Roundtrip Validation

## not working. we need at least one gate

```{r test01_roundtrip}
test_roundtrip(gs, export_path, test_dir, tolerance = 0.1)
```

---

# TEST 2: Single 1D Rectangle Gate (Linear)

**Goal:** Test simple 1D range gate on linear channel

## 2.1 Setup

```{r test02_setup}
test_name <- "TEST 2: 1D Rectangle Gate (Linear)"
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat(test_name, "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

test_dir <- file.path(BASE_TEST_DIR, "test02")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

ff <- create_test_fcs(n = 5000, seed = 2)
fcs_path <- file.path(test_dir, "sample02.fcs")
write.FCS(ff, fcs_path)

fs <- read.flowSet(fcs_path)
gs <- GatingSet(fs)

# Add 1D gate
gate <- rectangleGate(filterId = "FSC_filter", "FSC-A" = c(60000, 180000))
gs_pop_add(gs, gate, parent = "root")
recompute(gs)

cat("✓ GatingSet created\n")
cat("  Directory:", test_dir, "\n\n")

# Get counts for test info
stats <- gs_pop_get_count_fast(gs)

# Save test info
save_test_info(
  test_dir,
  test_name,
  description = "Single 1D rectangle gate (range gate) on FSC-A with linear transformation.",
  gate_info = c(
    "  Gate: FSC_filter",
    "  Type: rectangleGate (1D)",
    "  Parameter: FSC-A",
    "  Range: 60000 - 180000"
  ),
  transform_info = "All channels linear (no transformation)",
  expected_counts = c(
    sprintf("  root: %d", stats$Count[stats$Population == "root"]),
    sprintf("  FSC_filter: %d", stats$Count[stats$Population == "/FSC_filter"])
  )
)
```



## 2.4 Export

```{r test02_export}
export_path <- file.path(test_dir, "export.wsp")

success <- export_flowjo10_workspace(gs, export_path)


```

## 2.5 Inspect XML - Look for Gate

## 2.6 Roundtrip

```{r test02_roundtrip}
test_roundtrip(gs, export_path, test_dir, tolerance = 1.0)
```

---

# TEST 3: Single 2D Rectangle Gate (Linear)

**Goal:** Test 2D rectangle gate on linear channels

## 3.1 Setup

```{r test03_setup}
test_name <- "TEST 3: 2D Rectangle Gate (Linear)"
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat(test_name, "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

test_dir <- file.path(BASE_TEST_DIR, "test03")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

ff <- create_test_fcs(n = 5000, seed = 3)
fcs_path <- file.path(test_dir, "sample03.fcs")
write.FCS(ff, fcs_path)

fs <- read.flowSet(fcs_path)
gs <- GatingSet(fs)

# Add 2D gate
gate <- rectangleGate(
  filterId = "cells",
  "FSC-A" = c(60000, 180000),
  "SSC-A" = c(40000, 130000)
)
gs_pop_add(gs, gate, parent = "root")
recompute(gs)

cat("✓ GatingSet created\n")
cat("  Directory:", test_dir, "\n\n")

stats <- gs_pop_get_count_fast(gs)

save_test_info(
  test_dir,
  test_name,
  description = "Single 2D rectangle gate on FSC-A and SSC-A with linear transformation.",
  gate_info = c(
    "  Gate: cells",
    "  Type: rectangleGate (2D)",
    "  X-axis: FSC-A [60000, 180000]",
    "  Y-axis: SSC-A [40000, 130000]"
  ),
  transform_info = "All channels linear",
  expected_counts = c(
    sprintf("  root: %d", stats$Count[stats$Population == "root"]),
    sprintf("  cells: %d", stats$Count[stats$Population == "/cells"])
  )
)
```


## 3.4 Export

```{r test03_export}
export_path <- file.path(test_dir, "export.wsp")
success <- export_flowjo10_workspace(gs, export_path)

```

## 3.5 Inspect XML

## 3.6 Roundtrip

```{r test03_roundtrip}
test_roundtrip(gs, export_path, test_dir, tolerance = 1.0)
```

---

# TEST 4: Polygon Gate (Linear)

**Goal:** Test polygon gate on linear channels

## 4.1 Setup

```{r test04_setup}
test_name <- "TEST 4: Polygon Gate (Linear)"
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat(test_name, "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

test_dir <- file.path(BASE_TEST_DIR, "test04")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

ff <- create_test_fcs(n = 8000, seed = 4)
fcs_path <- file.path(test_dir, "sample04.fcs")
write.FCS(ff, fcs_path)

fs <- read.flowSet(fcs_path)
gs <- GatingSet(fs)

# Add polygon gate
gate <- polygonGate(
  filterId = "singlets",
  "FSC-A" = c(50000, 70000, 160000, 190000, 180000, 60000),
  "FSC-H" = c(45000, 40000, 140000, 180000, 190000, 80000)
)
gs_pop_add(gs, gate, parent = "root")
recompute(gs)

cat("✓ GatingSet created\n")
cat("  Directory:", test_dir, "\n\n")

stats <- gs_pop_get_count_fast(gs)

save_test_info(
  test_dir,
  test_name,
  description = "Polygon gate for singlet discrimination on FSC-A vs FSC-H with linear transformation.",
  gate_info = c(
    "  Gate: singlets",
    "  Type: polygonGate",
    "  X-axis: FSC-A",
    "  Y-axis: FSC-H",
    sprintf("  Vertices: %d", nrow(gate@boundaries))
  ),
  transform_info = "All channels linear",
  expected_counts = c(
    sprintf("  root: %d", stats$Count[stats$Population == "root"]),
    sprintf("  singlets: %d", stats$Count[stats$Population == "/singlets"])
  )
)
```



## 4.4 Export

```{r test04_export}
export_path <- file.path(test_dir, "export.wsp")
success <- export_flowjo10_workspace(gs, export_path)


```

## 4.5 Inspect XML


## 4.6 Roundtrip

```{r test04_roundtrip}
test_roundtrip(gs, export_path, test_dir, tolerance = 1.0)
```

---

# TEST 5: Ellipsoid Gate (Linear)

**Goal:** Test ellipsoid gate on linear channels

## 5.1 Setup

```{r test05_setup}
test_name <- "TEST 5: Ellipsoid Gate (Linear)"
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat(test_name, "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

test_dir <- file.path(BASE_TEST_DIR, "test05")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

ff <- create_test_fcs(n = 8000, seed = 5)
fcs_path <- file.path(test_dir, "sample05.fcs")
write.FCS(ff, fcs_path)

fs <- read.flowSet(fcs_path)
gs <- GatingSet(fs)

# Add ellipsoid gate
ellipse_cov <- matrix(c(1.5e9, 7e8, 7e8, 1e9), ncol = 2)
colnames(ellipse_cov) <- rownames(ellipse_cov) <- c("FSC-A", "SSC-A")

gate <- ellipsoidGate(
  filterId = "ellipse_cells",
  .gate = ellipse_cov,
  mean = c("FSC-A" = 120000, "SSC-A" = 80000),
  distance = 2
)
gs_pop_add(gs, gate, parent = "root")
recompute(gs)

cat("✓ GatingSet created\n")
cat("  Directory:", test_dir, "\n\n")

stats <- gs_pop_get_count_fast(gs)

save_test_info(
  test_dir,
  test_name,
  description = "Ellipsoid gate on FSC-A vs SSC-A with linear transformation.",
  gate_info = c(
    "  Gate: ellipse_cells",
    "  Type: ellipsoidGate",
    "  X-axis: FSC-A",
    "  Y-axis: SSC-A",
    "  Center: (120000, 80000)",
    "  Distance: 2"
  ),
  transform_info = "All channels linear",
  expected_counts = c(
    sprintf("  root: %d", stats$Count[stats$Population == "root"]),
    sprintf("  ellipse_cells: %d", stats$Count[stats$Population == "/ellipse_cells"])
  )
)
```


## 5.4 Export

```{r test05_export}
export_path <- file.path(test_dir, "export.wsp")
success <- export_flowjo10_workspace(gs, export_path)


```


## 5.6 Roundtrip

```{r test05_roundtrip}
test_roundtrip(gs_orig = gs, export_path = export_path, fcs_dir = test_dir, tolerance = 1.0)
```

---

# TEST 6: Biexponential Transform + 1D Gate

**Goal:** Test biex transformation with simple gate

## 6.1 Setup

```{r test06_setup}
test_name <- "TEST 6: Biexponential Transform"
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat(test_name, "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

test_dir <- file.path(BASE_TEST_DIR, "test06")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

ff <- create_test_fcs(n = 8000, seed = 6)
fcs_path <- file.path(test_dir, "sample06.fcs")
write.FCS(ff, fcs_path)

fs <- read.flowSet(fcs_path)
gs <- GatingSet(fs)

# Apply biex transform
biexp_trans <- flowjo_biexp_trans(
  channelRange = 4096,
  maxValue = 262144,
  pos = 4.5,
  neg = 0,
  widthBasis = -10
)
trans_list <- transformerList("FITC-A", biexp_trans)
gs <- transform(gs, trans_list)

# Add gate on transformed data
gate <- rectangleGate(filterId = "FITC_pos", "FITC-A" = c(1000, 3000))
gs_pop_add(gs, gate, parent = "root")
recompute(gs)

cat("✓ GatingSet created\n")
cat("  Directory:", test_dir, "\n\n")

stats <- gs_pop_get_count_fast(gs)

save_test_info(
  test_dir,
  test_name,
  description = "1D rectangle gate on FITC-A with biexponential transformation.",
  gate_info = c(
    "  Gate: FITC_pos",
    "  Type: rectangleGate (1D)",
    "  Parameter: FITC-A (transformed)",
    "  Range: 0.2 - 0.9 (transformed scale)"
  ),
  transform_info = c(
    "  FITC-A: Biexponential (FlowJo style)",
    "    channelRange: 4096",
    "    maxValue: 262144",
    "    pos: 4.5",
    "    neg: 0",
    "  Other channels: Linear"
  ),
  expected_counts = c(
    sprintf("  root: %d", stats$Count[stats$Population == "root"]),
    sprintf("  FITC_pos: %d", stats$Count[stats$Population == "/FITC_pos"])
  )
)
```


## 6.4 Export

```{r test06_export}
export_path <- file.path(test_dir, "export.wsp")
success <- export_flowjo10_workspace(gs, export_path)


```


## 6.6 Roundtrip

```{r test06_roundtrip}
test_roundtrip(gs_orig = gs, export_path = export_path, fcs_dir = test_dir, tolerance = 2.0)
```

---

# TEST 7: Log Transform + 1D Gate do we need this??

**Goal:** Test log transformation
there are many log transformations, logtGml2_trans is not working the same and has problems with the roundtrip
## 7.1 Setup

```{r test07_setup}
test_name <- "TEST 7: Log Transform"
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat(test_name, "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

test_dir <- file.path(BASE_TEST_DIR, "test07")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

ff <- create_test_fcs(n = 8000, seed = 7)
fcs_path <- file.path(test_dir, "sample07.fcs")
write.FCS(ff, fcs_path)

fs <- read.flowSet(fcs_path)
gs <- GatingSet(fs)

# Apply log transform
# log_trans <- flowWorkspace::logtGml2_trans( t=2000000)
log_trans = flowWorkspace::flowjo_log_trans(decade = 6, offset = 1, scale = 1, n=6)
trans_list <- transformerList("PE-A", log_trans)
gs <- transform(gs, trans_list)





# Add gate
# gate <- rectangleGate(filterId = "PE_pos", "PE-A" = c(100, 20000))
gate <- rectangleGate(filterId = "PE_pos", "PE-A" = c(0.4, 0.8))
gs_pop_add(gs, gate, parent = "root")
recompute(gs)

cat("✓ GatingSet created\n")
cat("  Directory:", test_dir, "\n\n")

stats <- gs_pop_get_count_fast(gs)

save_test_info(
  test_dir,
  test_name,
  description = "1D rectangle gate on PE-A with log transformation.",
  gate_info = c(
    "  Gate: PE_pos",
    "  Type: rectangleGate (1D)",
    "  Parameter: PE-A (log transformed)",
    "  Range: 0.3 - 0.95 (transformed scale)"
  ),
  transform_info = c(
    "  PE-A: Log (GML2)",
    "    t: 262144",
    "    m: 4.5",
    "  Other channels: Linear"
  ),
  expected_counts = c(
    sprintf("  root: %d", stats$Count[stats$Population == "root"]),
    sprintf("  PE_pos: %d", stats$Count[stats$Population == "/PE_pos"])
  )
)
```



## 7.4 Export

```{r test07_export}
export_path <- file.path(test_dir, "export.wsp")
success <- export_flowjo10_workspace(gs, export_path)


```


## 7.6 Roundtrip

```{r test07_roundtrip}
test_roundtrip(gs, export_path, test_dir, tolerance = 2.0)
```

---

# TEST 8: Arcsinh Transform + 1D Gate

**Goal:** Test arcsinh transformation

## 8.1 Setup

```{r test08_setup}
test_name <- "TEST 8: Arcsinh Transform"
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat(test_name, "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

test_dir <- file.path(BASE_TEST_DIR, "test08")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

ff <- create_test_fcs(n = 8000, seed = 8)
fcs_path <- file.path(test_dir, "sample08.fcs")
write.FCS(ff, fcs_path)

fs <- read.flowSet(fcs_path)
gs <- GatingSet(fs)

# Apply arcsinh transform
asinh_trans <- asinhtGml2_trans(T = 262144, M = 4.5, A = 0)
trans_list <- transformerList("APC-A", asinh_trans)
gs = transform(gs, trans_list)

# Add gate
gate <- rectangleGate(filterId = "APC_pos", "APC-A" = c(0.25, 0.92))
gs_pop_add(gs, gate, parent = "root")
recompute(gs)

cat("✓ GatingSet created\n")
cat("  Directory:", test_dir, "\n\n")

stats <- gs_pop_get_count_fast(gs)

save_test_info(
  test_dir,
  test_name,
  description = "1D rectangle gate on APC-A with arcsinh transformation.",
  gate_info = c(
    "  Gate: APC_pos",
    "  Type: rectangleGate (1D)",
    "  Parameter: APC-A (arcsinh transformed)",
    "  Range: 0.25 - 0.92 (transformed scale)"
  ),
  transform_info = c(
    "  APC-A: Arcsinh (GML2)",
    "    T: 262144",
    "    M: 4.5",
    "    A: 0",
    "  Other channels: Linear"
  ),
  expected_counts = c(
    sprintf("  root: %d", stats$Count[stats$Population == "root"]),
    sprintf("  APC_pos: %d", stats$Count[stats$Population == "/APC_pos"])
  )
)
```



## 8.4 Export



```{r test08_export}
export_path <- file.path(test_dir, "export.wsp")
success <- export_flowjo10_workspace(gs, export_path)


```



## 8.6 Roundtrip

```{r test08_roundtrip}
test_roundtrip(gs, export_path, test_dir, tolerance = 2.0)
```

---

# TEST 9: Hierarchical Gates (2 levels)

**Goal:** Test parent-child gate hierarchy

## 9.1 Setup

```{r test09_setup}
test_name <- "TEST 9: Hierarchical Gates"
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat(test_name, "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

test_dir <- file.path(BASE_TEST_DIR, "test09")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

ff <- create_test_fcs(n = 10000, seed = 9)
fcs_path <- file.path(test_dir, "sample09.fcs")
write.FCS(ff, fcs_path)

fs <- read.flowSet(fcs_path)
gs <- GatingSet(fs)

# Level 1: Parent gate
parent_gate <- rectangleGate(
  filterId = "cells",
  "FSC-A" = c(60000, 190000),
  "SSC-A" = c(40000, 140000)
)
gs_pop_add(gs, parent_gate, parent = "root")
recompute(gs)

# Level 2: Child gate
child_gate <- rectangleGate(
  filterId = "singlets",
  "FSC-A" = c(70000, 180000),
  "FSC-H" = c(60000, 170000)
)
gs_pop_add(gs, child_gate, parent = "/cells")
recompute(gs)

cat("✓ GatingSet created\n")
cat("  Directory:", test_dir, "\n\n")

stats <- gs_pop_get_count_fast(gs)

save_test_info(
  test_dir,
  test_name,
  description = "Two-level hierarchical gating: cells → singlets",
  gate_info = c(
    "  Level 1 - cells:",
    "    Type: rectangleGate (2D)",
    "    Parameters: FSC-A, SSC-A",
    "    Parent: root",
    "  Level 2 - singlets:",
    "    Type: rectangleGate (2D)",
    "    Parameters: FSC-A, FSC-H",
    "    Parent: cells"
  ),
  transform_info = "All channels linear",
  expected_counts = c(
    sprintf("  root: %d", stats$Count[stats$Population == "root"][1]),
    sprintf("  cells: %d", stats$Count[stats$Population == "/cells"][1]),
    sprintf("  cells/singlets: %d", stats$Count[stats$Population == "/cells/singlets"][1])
  )
)
```



## 9.4 Export

```{r test09_export}
export_path <- file.path(test_dir, "export.wsp")
success <- export_flowjo10_workspace(gs, export_path)


```


## 9.6 Roundtrip

```{r test09_roundtrip}
test_roundtrip(gs, export_path, test_dir, tolerance = 1.0)
```

---

# TEST 10: Boolean Gate (AND)

**Goal:** Test boolean AND operation

## 10.1 Setup

```{r test10_setup}
test_name <- "TEST 10: Boolean AND Gate"
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat(test_name, "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

test_dir <- file.path(BASE_TEST_DIR, "test10")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

ff <- create_test_fcs(n = 8000, seed = 10)
fcs_path <- file.path(test_dir, "sample10.fcs")
write.FCS(ff, fcs_path)

fs <- read.flowSet(fcs_path)
gs <- GatingSet(fs)

# Add two independent gates
gate1 <- rectangleGate(filterId = "FSC_gate", "FSC-A" = c(70000, 200000))
gs_pop_add(gs, gate1, parent = "root")

gate2 <- rectangleGate(filterId = "SSC_gate", "SSC-A" = c(50000, 130000))
gs_pop_add(gs, gate2, parent = "root")

recompute(gs)

# Boolean AND
bool_gate <- booleanFilter(`FSC_gate&SSC_gate`, filterId = "both")
gs_pop_add(gs, bool_gate, parent = "root")
bool_gate2 <- booleanFilter(`FSC_gate|SSC_gate`, filterId = "bothOR")
gs_pop_add(gs, bool_gate2, parent = "root")
bool_gate2 <- booleanFilter(`!both`, filterId = "bothNOT")
gs_pop_add(gs, bool_gate2, parent = "root")

recompute(gs)

cat("✓ GatingSet created\n")
cat("  Directory:", test_dir, "\n\n")

stats <- gs_pop_get_count_fast(gs)

save_test_info(
  test_dir,
  test_name,
  description = "Boolean AND gate combining two independent rectangle gates",
  gate_info = c(
    "  Gate 1 - FSC_gate:",
    "    Type: rectangleGate (1D)",
    "    Parameter: FSC-A",
    "  Gate 2 - SSC_gate:",
    "    Type: rectangleGate (1D)",
    "    Parameter: SSC-A",
    "  Boolean - both:",
    "    Expression: FSC_gate AND SSC_gate",
    "    Type: booleanFilter"
  ),
  transform_info = "All channels linear",
  expected_counts = c(
    sprintf("  root: %d", stats$Count[stats$Population == "root"][1]),
    sprintf("  FSC_gate: %d", stats$Count[stats$Population == "/FSC_gate"][1]),
    sprintf("  SSC_gate: %d", stats$Count[stats$Population == "/SSC_gate"][1]),
    sprintf("  both: %d", stats$Count[stats$Population == "/both"][1])
  )
)
```



## 10.4 Export

```{r test10_export}
export_path <- file.path(test_dir, "export.wsp")
success <- export_flowjo10_workspace(gs, export_path)


```


## 10.6 Roundtrip

```{r test10_roundtrip}
test_roundtrip(gs, export_path, test_dir, tolerance = 1.0)
```

---

# TEST 11: CyDa output

includes asinh transformations and some gates?

## 11.1 Setup

```{r test11_setup}
gs = load_gs("/pasteur/helix/projects/LabExMI/USERS/bernd/cytometry/analysis/adaptiveFunctional/channel_rename/keepChannels/indexed/rmMargins/noDoublets/cyEt/transformed/_data/GS")

test_name <- "TEST 11: CyDa"
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat(test_name, "\n")
cat(paste(rep("=", 70), collapse = ""), "\n\n")

test_dir <- file.path(BASE_TEST_DIR, "test11")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

for (i in seq_along(gs)) {
  gh <- gs[[i]]
  
  # Get the root population (all events, with transformations applied)
  fr <- gh_pop_get_data(gh, "root")
  # Save to FCS
  write.FCS(fr, 
            file = file.path(test_dir, paste0(sampleNames(gs)[i], ".fcs")))
}

# Add two independent gates
# Add 2D gate
gate1 <- rectangleGate(
  filterId = "test1",
  "Live.Dead.A" = c(-13, -10.4),
  "CD3.A" = c(-9.8, -0.5)
)
gs_pop_add(gs, gate1, parent = "root")

gate2 <- rectangleGate(
  filterId = "test2",
  "Live.Dead.A" = c(-10.2, -4.6),
  "CD3.A" = c(-3.3, -0.1)
)
gs_pop_add(gs, gate2, parent = "root")

# Add polygon gate
gate <- polygonGate(
  filterId = "singlets",
  "Live.Dead.A" = c(-6.39, -3.8, -3.5, -7.9, -8.9),
  "CD3.A" = c(0.59, -0.15, -4.24, -9.04, -7.39)
)
gs_pop_add(gs, gate, parent = "root")
recompute(gs)



```


## 11.4 Export

DEBUG

```{r test11_export}
export_path <- file.path(test_dir, "export.wsp")
success <- export_flowjo10_workspace(gs, export_path)


```


## 11.6 Roundtrip

```{r test11_roundtrip}
test_roundtrip(gs, export_path, test_dir, tolerance = 2.0)
```

---

# Summary: Files for FlowJo Testing

```{r summary}
cat("\n")
cat(paste(rep("=", 80), collapse = ""), "\n")
cat("SUMMARY: FILES FOR FLOWJO VALIDATION\n")
cat(paste(rep("=", 80), collapse = ""), "\n\n")

cat("Base directory:", BASE_TEST_DIR, "\n")
cat("Absolute path:", normalizePath(BASE_TEST_DIR, mustWork = FALSE), "\n\n")

test_dirs <- list.dirs(BASE_TEST_DIR, recursive = FALSE)
test_dirs <- test_dirs[basename(test_dirs) != ""]

if (length(test_dirs) > 0) {
  cat("Test directories created:\n\n")
  
  for (td in sort(test_dirs)) {
    test_num <- basename(td)
    wsp_file <- file.path(td, "export.wsp")
    info_file <- file.path(td, "test_info.txt")
    
    if (file.exists(wsp_file)) {
      cat(sprintf("  [%s]\n", test_num))
      cat(sprintf("    Workspace: %s\n", wsp_file))
      
      if (file.exists(info_file)) {
        # Read first line (test name)
        info_lines <- readLines(info_file, n = 3)
        if (length(info_lines) > 0) {
          cat(sprintf("    %s\n", info_lines[1]))
        }
      }
      cat("\n")
    }
  }
  
  cat(paste(rep("-", 80), collapse = ""), "\n")
  cat("TO TEST IN FLOWJO:\n")
  cat("  1. Open FlowJo v10\n")
  cat("  2. File → Import → Workspace\n")
  cat("  3. Navigate to test directory above\n")
  cat("  4. Select a test's export.wsp file\n")
  cat("  5. Verify gates and statistics\n")
  cat("  6. Read test_info.txt for expected results\n")
  cat(paste(rep("=", 80), collapse = ""), "\n")
} else {
  cat("No test directories found. Run the test chunks above first.\n")
}
```

---

# Session Info

```{r session_info}
sessionInfo()
```


