---
title: "Compensation Workflow with CyFj11"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Compensation Workflow with CyFj11}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{CyFj11}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CyFj11)
library(flowWorkspace)
library(flowCore)
library(ggcyto)
```

## Introduction

This vignette walks through a complete compensation-focused workflow using
`CyFj11`. It demonstrates how to:

1. Read an FCS file and inspect its spillover matrix.
2. Modify the spillover matrix and write a new FCS file.
3. Parse a FlowJo v11 workspace.
4. Convert the workspace into a `GatingSet`.
5. Merge per-sample `GatingSet` objects and extract population counts.
6. Visualize the data with `ggcyto`.

The code below mirrors the workflow captured in the package's
`test.compensation.R` development script. Example file paths point to the
`flowjo_export_tests/test14/` directory, which is available in the development
repository but not shipped with the package. Set `eval = TRUE` only when those
files are present, or adapt the paths to your own data.

## Inspecting and modifying compensation

Start by reading an FCS file and looking at its compensation-related keywords.

```{r inspect-fcs, eval=FALSE}
# Path to an FCS file referenced by a FlowJo v11 workspace
fcs_path <- "flowjo_export_tests/test14/68983.fcs"

ff <- flowCore::read.FCS(fcs_path)
kw <- flowCore::keyword(ff)
kw[grep("SPILL|COMP", names(kw), ignore.case = TRUE)]
```

Extract the original spillover matrix, scale it (for example by 100), and
attach the modified matrix to a new `flowFrame`.

```{r modify-spillover, eval=FALSE}
spill_orig <- flowCore::keyword(ff, "SPILL")[[1]]
cat("Original spillover matrix:\n")
print(spill_orig)

spill_new <- spill_orig * 100
flowCore::keyword(ff)[["SPILL"]] <- spill_new

modified_fcs_path <- "flowjo_export_tests/test14/68983_modified.fcs"
flowCore::write.FCS(ff, modified_fcs_path)
```

Verify that the resaved file contains the updated spillover matrix.

```{r verify-spillover, eval=FALSE}
ff_check <- flowCore::read.FCS(modified_fcs_path, transformation = FALSE)
spill_check <- flowCore::keyword(ff_check, "SPILL")[[1]]
cat("Spillover in resaved file:\n")
print(spill_check)
```

## Parsing the FlowJo v11 workspace

Next, parse the FlowJo v11 workspace that references the FCS file.

```{r parse-workspace, eval=FALSE}
fj_path <- "flowjo_export_tests/test14/test14.compensation.flowjo"

ws <- tryCatch(
  read_flowjo11_workspace(fj_path),
  error = function(e) {
    warning("Failed to parse ", fj_path, ": ", e$message)
    return(NULL)
  }
)
if (is.null(ws)) stop("Workspace parsing failed")
```

## Converting to a GatingSet

`fj11_to_gatingset()` returns a named list of `GatingSet` objects, one per
sample. Because the FCS files live next to the workspace, we use the workspace
directory as the search path.

```{r convert-gatingset, eval=FALSE}
fcs_search_path <- dirname(fj_path)

gs_list <- tryCatch(
  fj11_to_gatingset(
    fj11_workspace     = ws,
    group_name         = 1,
    path               = fcs_search_path,
    execute            = TRUE,
    stop_on_multiple   = FALSE,
    include_empty_tree = FALSE
  ),
  error = function(e) {
    warning("Failed to convert ", fj_path, " to GatingSet: ", e$message)
    return(NULL)
  }
)
if (is.null(gs_list) || length(gs_list) == 0) stop("GatingSet conversion failed")
```

## Merging and summarizing

Merge the per-sample list into a single `GatingSet` for downstream analysis and
extract population counts.

```{r merge-counts, eval=FALSE}
gs <- flowWorkspace::merge_list_to_gs(gs_list)

# Population counts from the GatingSet
stats <- flowWorkspace::gs_pop_get_stats(gs)
head(stats)
```

## Visualization

After conversion, plot a sample. When transformations are applied by
`fj11_to_gatingset()`, `limits = "data"` usually gives a clearer view of the
gated populations than the raw 0-262,144 linear range.

```{r plot, eval=FALSE}
autoplot(gs[[1]], bins = 264) +
  ggcyto::ggcyto_par_set(limits = "data")
```

## Complete script

Putting the pieces together:

```{r complete, eval=FALSE}
library(CyFj11)
library(flowCore)
library(flowWorkspace)

fcs_path   <- "flowjo_export_tests/test14/68983.fcs"
fj_path    <- "flowjo_export_tests/test14/test14.compensation.flowjo"

# 1. Modify compensation
ff        <- flowCore::read.FCS(fcs_path)
spill_new <- flowCore::keyword(ff, "SPILL")[[1]] * 100
flowCore::keyword(ff)[["SPILL"]] <- spill_new
flowCore::write.FCS(ff, "flowjo_export_tests/test14/68983_modified.fcs")

# 2. Parse workspace
ws <- read_flowjo11_workspace(fj_path)

# 3. Convert to GatingSet
gs_list <- fj11_to_gatingset(
  fj11_workspace = ws,
  group_name     = 1,
  path           = dirname(fj_path),
  execute        = TRUE
)

# 4. Merge and inspect
gs <- flowWorkspace::merge_list_to_gs(gs_list)
autoplot(gs[[1]], bins = 264) + ggcyto::ggcyto_par_set(limits = "data")
```

## See also

* `vignette("getting-started")` for the basic import/export workflow.
* `?fj11_to_gatingset` for all conversion options, including the
  `use_transformed_coords` and `transform` arguments that control whether gate
  coordinates remain in FlowJo display space.
