---
title: "Compensation and Round-Trip Workflow with CyFj11"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Compensation and Round-Trip Workflow with CyFj11}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{CyFj11}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Only evaluate data-dependent chunks when FlowSOM is available.
has_flowsom <- requireNamespace("FlowSOM", quietly = TRUE)
```

```{r setup}
library(CyFj11)
library(flowCore)
library(flowWorkspace)
library(ggcyto)
```

## Introduction

This vignette walks through a complete round-trip workflow using public example
data from the Bioconductor `FlowSOM` package. The workflow demonstrates how to:

1. Locate the public FlowSOM example FCS file and v10 workspace.
2. Import the v10 workspace into a `GatingSet` with `CytoML`.
3. Export the `GatingSet` back to a FlowJo v10 `.wsp` file with `CyFj11`.
4. Re-import the exported workspace and compare population counts.
5. Verify that the compensation (`SPILL`) matrix survives the round trip.
6. Visualize the result with `ggcyto`.

All data are taken unchanged from `FlowSOM`; no compensation matrix is modified.

## Data availability

The example data are bundled with the `FlowSOM` package. If it is not installed,
you can install it from Bioconductor:

```{r install-flowsom, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("FlowSOM")
```

Locate the files:

```{r locate-flowsom, eval=has_flowsom}
flowsom_fcs <- system.file("extdata", "68983.fcs", package = "FlowSOM")
flowsom_wsp <- system.file("extdata", "gating.wsp", package = "FlowSOM")

stopifnot(file.exists(flowsom_fcs), file.exists(flowsom_wsp))
cat("FCS:", flowsom_fcs, "\n")
cat("WSP:", flowsom_wsp, "\n")
```

Copy the FCS file into a clean temporary directory so the workspace can find it
by its original name:

```{r prepare-temp, eval=has_flowsom}
test_dir <- tempfile("cyfj11_flowsom_")
dir.create(test_dir, showWarnings = FALSE)
file.copy(flowsom_fcs, file.path(test_dir, "68983.fcs"), overwrite = TRUE)
cat("Working directory:", test_dir, "\n")
```

## Import the FlowSOM v10 workspace

Use `CytoML` to read the FlowJo v10 XML workspace and build a `GatingSet`.

```{r import-v10, eval=has_flowsom}
library(CytoML)

ws <- CytoML::open_flowjo_xml(flowsom_wsp)
gs <- CytoML::flowjo_to_gatingset(
  ws,
  path                = test_dir,
  name                = "All Samples",
  subset              = "68983.fcs",
  sample_names_from   = "sampleID"
)

cat("Samples:", length(gs), "\n")
cat("Populations:\n")
print(flowWorkspace::gs_get_pop_paths(gs))
```

Inspect the population counts:

```{r counts-before, eval=has_flowsom}
stats_before <- flowWorkspace::gs_pop_get_count_fast(gs)
head(stats_before[, c("Population", "Count")])
```

## Export the GatingSet to FlowJo v10 with CyFj11

Now export the same `GatingSet` to a new FlowJo v10 `.wsp` file using
`CyFj11::export_flowjo10_workspace()`.

```{r export-v10, eval=has_flowsom}
export_path <- file.path(test_dir, "test14_export.wsp")
CyFj11::export_flowjo10_workspace(gs, export_path)

stopifnot(file.exists(export_path))
cat("Exported workspace:", export_path, "\n")
```

## Re-import and validate

Read the exported workspace back with `CytoML` and compare the population
counts. They should match the original import.

```{r reimport, eval=has_flowsom}
ws2 <- CytoML::open_flowjo_xml(export_path)
gs2 <- CytoML::flowjo_to_gatingset(
  ws2,
  path              = test_dir,
  name              = "All Samples",
  subset            = "68983.fcs",
  sample_names_from = "sampleID"
)

stats_after <- flowWorkspace::gs_pop_get_count_fast(gs2)

all_equal <- all(stats_before$Count == stats_after$Count)
cat("Population counts identical:", all_equal, "\n")

if (!all_equal) {
  diff_idx <- which(stats_before$Count != stats_after$Count)
  print(cbind(
    Population = stats_before$Population[diff_idx],
    Before     = stats_before$Count[diff_idx],
    After      = stats_after$Count[diff_idx]
  ))
}
```

## Verify compensation matrix survival

A key part of the round trip is that the FCS `SPILL` matrix is preserved in the
exported FlowJo workspace. We compare the original FCS `SPILL` keyword with the
`spilloverMatrix` written into the exported `.wsp` file.

```{r verify-compensation, eval=has_flowsom}
orig_spill <- flowCore::keyword(flowCore::read.FCS(flowsom_fcs), "SPILL")[[1]]

# Extract SPILL keyword from the exported workspace's Sample keywords.
wsp_lines <- readLines(export_path)
spill_line <- wsp_lines[grep('name="SPILL"', wsp_lines)]
flat <- sub('.*value="', "", spill_line)
flat <- sub('" */>.*', "", flat)
parts <- strsplit(flat, ",")[[1]]
n <- as.integer(parts[1])
exp_names <- parts[2:(n + 1)]
exp_vals <- as.numeric(parts[(n + 2):(n + 1 + n * n)])
exp_spill <- matrix(exp_vals, nrow = n, ncol = n,
                    dimnames = list(exp_names, exp_names))

# Exported matrix is on the same 0-1 proportion scale as the original FCS keyword.
exp_spill_ordered <- exp_spill[colnames(orig_spill), colnames(orig_spill)]
max_diff <- max(abs(exp_spill_ordered - orig_spill))
cat("SPILL matrix dimension:", n, "x", n, "\n")
cat("Max absolute element-wise difference:", max_diff, "\n")
stopifnot(max_diff < 1e-6)
```

## Visualization

Plot a gate from the imported `GatingSet`. Because the original FlowSOM
workspace uses linear scaling, the default axes are appropriate here. For
biexponential channels you may prefer `ggcyto_par_set(limits = "data")`.

```{r plot, eval=has_flowsom, fig.width=7, fig.height=6}
pops <- flowWorkspace::gs_get_pop_paths(gs)
plot_pop <- if (length(pops) > 1) pops[2] else pops[1]

ggcyto::autoplot(gs[[1]], plot_pop, bins = 128)
```

## Complete script

```{r complete, eval=FALSE}
library(CyFj11)
library(flowCore)
library(flowWorkspace)
library(CytoML)
library(ggcyto)

# 1. Locate FlowSOM example data
flowsom_fcs <- system.file("extdata", "68983.fcs", package = "FlowSOM")
flowsom_wsp <- system.file("extdata", "gating.wsp", package = "FlowSOM")

# 2. Prepare a temporary directory with the FCS file
test_dir <- tempfile("cyfj11_flowsom_")
dir.create(test_dir)
file.copy(flowsom_fcs, file.path(test_dir, "68983.fcs"))

# 3. Import the FlowJo v10 workspace
ws  <- CytoML::open_flowjo_xml(flowsom_wsp)
gs  <- CytoML::flowjo_to_gatingset(ws, path = test_dir, name = "All Samples",
                                    subset = "68983.fcs",
                                    sample_names_from = "sampleID")

# 4. Export with CyFj11
export_path <- file.path(test_dir, "test14_export.wsp")
CyFj11::export_flowjo10_workspace(gs[[1]], export_path)

# 5. Re-import and compare counts
ws2 <- CytoML::open_flowjo_xml(export_path)
gs2 <- CytoML::flowjo_to_gatingset(ws2, path = test_dir, name = "All Samples",
                                    subset = "68983.fcs",
                                    sample_names_from = "sampleID")
all(flowWorkspace::gs_pop_get_count_fast(gs)$Count ==
    flowWorkspace::gs_pop_get_count_fast(gs2)$Count)

# 6. Plot
pops <- flowWorkspace::gs_get_pop_paths(gs)
ggcyto::autoplot(gs[[1]], pops[2], bins = 128)
```

## See also

* `vignette("getting-started")` for the basic FlowJo v11 import workflow.
* `?export_flowjo10_workspace` for details on exporting a `GatingSet` to FlowJo
  v10 format.
* `?fj11_to_gatingset` for importing FlowJo v11 workspaces.
