---
title: "Exporting to FlowJo v10 with CyFj11"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting to FlowJo v10 with CyFj11}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CyFj11)
```

## Introduction

This vignette focuses on CyFj11's FlowJo v10 export functionality, which enables researchers to export R-based gating analyses back to the FlowJo environment. While CyFj11 provides **asymmetric format conversion**—full support for importing FlowJo v11 workspaces into R GatingSet objects, but only v10 format export—this design choice is technically justified and practically viable.

The asymmetric conversion design reflects the technical realities of developing against proprietary formats without vendor documentation:

1. **FlowJo v11 Complexity**: The v11 JSON format uses extensive UUID cross-referencing between components, with embedded transformations and a complex schema.

2. **Practical Solution**: The v10 XML format follows the documented **Gating-ML 2.0 standard**, making it feasible to generate valid workspaces through schema validation. The XML structure is transparent and amenable to debugging.

## Basic Export Workflow

The export process converts a GatingSet object to a FlowJo v10 workspace file:

```{r basic-export, eval=FALSE}
# Assuming you have a GatingSet object from previous analysis
# gs <- your_analysis_results

# Export to FlowJo v10 format
export_path <- "path/to/output/workspace.wsp"
success <- export_flowjo10_workspace(gs, export_path)

if (success) {
  cat("Successfully exported to:", export_path, "\n")
} else {
  cat("Export failed.\n")
}
```

## Supported Gate Types and Transformations

The export function handles standard flowCore gate types and common transformations:

### Gate Types

```{r gate-types, eval=FALSE}
# Supported gate types include:
# - rectangleGate: Converted to <gating:RectangleGate> with dimension-specific min/max bounds
# - polygonGate: Converted to <gating:PolygonGate> with vertex coordinates
# - ellipsoidGate: Converted using eigenvalue decomposition to derive rotation angles and axis lengths
# - booleanFilter: Converted to <gating:BooleanGate> with expression preservation

# Check what gate types are in your GatingSet
gate_types <- sapply(gs_pop_get_gate(gs, gs_get_pop_paths(gs)[-1]), class)
table(unlist(gate_types))
```

### Transformations

Various transformation types are supported:

```{r transformations, eval=FALSE}
# Supported transformations include:
# - Biexponential (biexp): Maps to FlowJo's biexponential transform parameters (T, W, M, A)
# - Linear: Preserved with min/max range specifications
# - Log: Converted with decade and offset parameters
# - Logicle: Mapped to FlowJo's logicle implementation
# - Arcsinh (fasinh): Converted with appropriate scaling parameters

# Check transformations in your GatingSet
sample_name <- sampleNames(gs)[1]
trans <- gh_get_transformations(gs[[sample_name]])
names(trans)
```

## FlowJo Compatibility

The exported v10 workspaces maintain broad compatibility with FlowJo v11:

### Importing Exported Workspaces

```{r flowjo-import, eval=FALSE}
# FlowJo v11 can import v10 workspaces without significant information loss:
# 1. Open FlowJo v11
# 2. File → Import → Workspace
# 3. Select the exported .wsp file
# 4. All gates, populations, and statistics should be preserved
```

## Limitations

While the v10 export format provides substantial functionality, it has some limitations compared to native v11 features:

1. **Feature Set**: Researchers requiring v11-specific features (advanced derived parameters, hierarchical Boolean logic, integrated statistical overlays) cannot directly export these from R to FlowJo.

2. **Rich Text Formatting**: Some annotation text formatting may be lost during conversion.

Despite these limitations, the v10 interchange format remains supported by current FlowJo versions and preserves essential gating information including rectangle, polygon, ellipsoid, and Boolean gates with their coordinate transformations.

## Best Practices

Follow these best practices for successful exports:

1. **Validate Before Export**: Check your GatingSet for unsupported features
2. **Test with Small Workspaces**: Validate the process with simpler workspaces first
3. **Document Transformations**: Keep records of any custom transformations used
4. **Verify Results**: Always check exported files in FlowJo to ensure correctness

## Next Steps

For related topics, see:

* `vignette("getting-started")` for basic workflow