---
title: "Getting Started with CyFj11"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with CyFj11}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(CyFj11)
```

## Introduction

CyFj11 is an R package designed to bridge the gap between FlowJo v11 workspaces and the R flowWorkspace ecosystem. It provides **asymmetric format conversion** capabilities:

* **Import**: Full support for parsing FlowJo v11 workspaces (.flowjo files) into R GatingSet objects
* **Export**: Generation of FlowJo v10 workspace files (.wsp) from GatingSet objects with broad compatibility

This asymmetric design reflects the technical realities of developing against proprietary formats without vendor documentation. While FlowJo v11 uses a complex JSON schema with extensive UUID cross-referencing that makes direct export impractical, the v10 XML format follows the documented Gating-ML 2.0 standard, making it feasible to generate valid workspaces.

## Installation

To install CyFj11, ensure you have the required dependencies:

```{r install, eval=FALSE}
# Install from GitHub (or CRAN if available)
# devtools::install_github("baj12/CyFj11")

# Load the package
library(CyFj11)
```

## Basic Workflow

The typical workflow with CyFj11 involves three steps:

1. Import a FlowJo v11 workspace
2. Analyze the data in R
3. Export results back to FlowJo (v10 format)

Let's walk through this process with a simple example.

### Importing a FlowJo v11 Workspace

First, we'll read a FlowJo v11 workspace file:

```{r read-workspace, eval=FALSE}
# Set the path to your FlowJo workspace file
workspace_path <- system.file("extdata", "test.data.flowjo", package = "CyFj11")

# Read the FlowJo v11 workspace
ws <- read_flowjo11_workspace(workspace_path)

# Examine the workspace structure
print(names(ws))
```

The `read_flowjo11_workspace()` function parses the ZIP-based FlowJo v11 format and extracts the JSON data structures. The returned object contains all the information from the workspace, including groups, samples, gates, and populations.

### Converting to GatingSet

Next, we convert the parsed workspace to a GatingSet object, which is the standard format for flow cytometry data in R:

```{r convert-gs, eval=FALSE}
workspace_fcs <- system.file("extdata", "", package = "CyFj11")
# You'll need to specify the path to your FCS files
gs <- fj11_to_gatingset(
  fj11_workspace = ws,
  group_name = 1,
  path = workspace_fcs
)


# Check the number of samples
length(gs)

# View sample names
sampleNames(gs[[1]])
gs = merge_list_to_gs(gs)
```

Note:

If you don't know the group name you can leave it empty and then you will be asked :

Choose which group of samples to import: 

1: Acquired Data
2: Compensation Data
3: Experiment Data

Note:
fj11_to_gatingset is returning a list of gatingset objects. The can be combined to a gatingset with multiple sample by using merge_list_to_gs, but this might overwrite transformation parameters or other sample specific settings from FlowJo11

Note that you'll need to provide the path to your FCS files. The `fj11_to_gatingset()` function will attempt to locate the FCS files referenced in the workspace.

### Examining the GatingSet

Once we have a GatingSet, we can examine its structure:

```{r examine-gs, eval=FALSE}
# View the population hierarchy
gs_get_pop_paths(gs)

# Get population statistics
stats <- gs_pop_get_stats(gs)
head(stats)

plot(gs)
autoplot(gs, "cells", bins = 80)
autoplot(gs, "cd8Pos", bins = 80)

```

### Exporting Back to FlowJo

After performing analysis in R, you can export your results back to FlowJo v10 format:

```{r export, eval=FALSE}
# Export to FlowJo v10 format
export_path <- tempfile(fileext = ".wsp")
export_flowjo10_workspace(gs[[1]], export_path)

# Verify the export was successful
file.exists(export_path)
```
Note:

warnings about invalid transformations are normal, as we don't have any transformations set in FlowJo 11 and the gatingset.

The exported workspace can be opened in FlowJo v11, which maintains backward compatibility with v10 workspaces.

## Next Steps

For more detailed information about exporting to FlowJo, see `vignette("exporting-to-flowjo")`.

For function-specific documentation, use the help system:
```{r help, eval=FALSE}
?read_flowjo11_workspace
?fj11_to_gatingset
?export_flowjo10_workspace
